What to Expect with Life Expectancy?

### 2024-12-04 paper/paper.qmd
Add files via upload

---
title: "What to Expect with Life Expectancy?"
subtitle: "Examining How Status, Healthcare Expenditure, Education, and Medical Infrastructure Influence Life Expectancy Across Developed and Developing Regions"
author: 
  - Manjun Zhu
thanks: "Code and data are available at: [https://github.com/Karrrrmen/Life_Expectancy](https://github.com/Karrrrmen/Life_Expectancy)."
date: today
date-format: long
abstract: "Life expectancy has seen unprecedented growth in recent decades, driven by advancements in healthcare, improved access to medical services, and reduced child mortality. However, this growth has not been evenly distributed across the globe, with wealthier nations benefiting disproportionately. This research examines the impact of a nation’s economic standing and health conditions on life expectancy, exploring nuanced variables beyond those traditionally studied. By developing a linear model, the analysis identifies key relationships between economic and healthcare indicators and their influence on population longevity. The findings aim to inform health policy and economic planning, providing actionable insights to address disparities and promote equitable growth in life expectancy worldwide."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

# load libraries
library(dplyr)
library(tidyverse)
library(here)
library(ggplot2)
library(knitr)
library(kableExtra)
library(modelsummary)
library(ggbeeswarm)
library(ggridges)

# load data
data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))

@@

Life expectancy has witnessed remarkable growth over the past century, driven by advancements in healthcare, improved access to medical services, and reduced child mortality rates. However, this progress has not been distributed equitably across the globe. While nations with robust economies and advanced healthcare systems have achieved significant gains, others, particularly in underdeveloped regions, continue to face barriers such as inadequate infrastructure, persistent diseases, and economic instability. Despite existing research on life expectancy determinants, gaps remain in understanding the nuanced relationships between economic standing, healthcare access, and other critical variables.

This study aims to examine the factors influencing life expectancy across different nations by developing a linear model that incorporates both healthcare and economic variables. The response variable in the analysis is life expectancy, while the predictors include national status (Developed vs. Developing), diphtheria immunization coverage, government healthcare expenditure as a percentage of total expenditure, and the number of doctors and nurses per 10,000 people. Additionally, schooling is considered as an important educational factor. By quantifying the effects of these predictors, the study seeks to shed light on disparities in life expectancy between developed and developing nations, and to identify key factors contributing to these differences. Through this analysis, we aim to contribute valuable insights for policy development focused on improving healthcare and economic outcomes.

Previous studies provide valuable context for this research. One study [@Garon2015] examined the long-term effects of diphtheria on survivors, revealing that paralyzed individuals experience greater reductions in life expectancy, which underscores the importance of healthcare infrastructure in mitigating preventable diseases in underdeveloped regions. Another study extended this understanding by exploring the role of the Human Development Index (HDI), emphasizing the influence of public health initiatives, economic conditions, and education on life expectancy in low and medium-HDI nations [@Girum2018]. Besides that, advanced econometric techniques, including panel cointegration analysis, have demonstrated a positive relationship between health care expenditure per capita and life expectancy, providing a robust foundation for integrating economic variables into this study [@He2020]. Additionally, a finding [@Hanushek2016] underscored the importance of combining both educational investments and government support in improving life expectancy, particularly in developing regions. The strong correlation between education and health suggests that higher levels of education improve overall health, reducing mortality rates and extending life expectancy.

The analysis revealed that life expectancy in developed countries is generally higher and more stable, attributed to stronger healthcare systems, consistent access to medical services, and robust economic stability. These factors are typically supported by well-established public health policies. In contrast, life expectancy in developing countries demonstrates a broader range of variation, though with a noticeable upward trend. This improvement is driven by diverse factors like better healthcare access, economic development, and increasing resource availability, but the inconsistency across regions highlights disparities in infrastructure. The faster growth trajectory in these nations indicates progress, yet challenges remain in achieving stable health outcomes. These findings suggest that policymakers in developing countries should prioritize healthcare system strengthening, resource access, and reducing regional inequalities, while in developed countries, maintaining consistent healthcare access and supporting sustained economic growth should remain priorities.

@@

This study utilizes two primary datasets related to global health and life expectancy. The first dataset, World Health Statistics 2020 Complete, sourced from Kaggle [@world_health_statistics], provides comprehensive global health data, including variables such as GDP, life expectancy, mortality rates, and access to healthcare. The second dataset, Life Expectancy (WHO), also from Kaggle [@life_expectancy_who], focuses specifically on life expectancy trends and factors, offering a detailed breakdown by country, year, and socioeconomic variables.

@@

We performed data cleaning and analysis using R [@citeR], incorporating a range of packages for efficient data manipulation, visualization, and statistical modeling, leveraging packages such as **`tidyverse`** for data manipulation and visualization [@tidyverse], **`dplyr`** for streamlined data wrangling [@dplyr], **`ggplot2`** for creating high-quality plots [@ggplot2], **`ggridges`** for displaying the distribution of variables across different categories with ridge plots [@ggridges], **`here`** for managing file paths [@here], **`lubridate`** for handling date-related variables [@lubridate], **`kableExtra`** for generating polished tables [@kableExtra], **`modelsummary`** for summarizing model results [@modelsummary].

@@

The measurement process refers to how real-world phenomena—such as healthcare access, economic indicators, or demographic factors—are translated into numerical entries in the dataset. Each entry corresponds to a specific country and year, capturing variables related to life expectancy and associated determinants. Below is a description of key variables in the dataset:

@@

**Health Expenditure (`TotalExpenditure`)**: This represents the general government expenditure on health as a percentage of total government expenditure. It is reported by national governments or international organizations and recorded as a percentage in the dataset.

@@

```{r}
#| label: fig-response
#| fig-cap: "Distribution of life expectancy across different Status groups (Developed vs. Developing), highlighting the distinct patterns in their density. The mode of the density for life expectancy in developed countries tends to be around 81 years with a narrower variance, whereas in developing countries, the mode is approximately 74 years with a wider variance, reflecting disparities in health outcomes."
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(x = LifeExpectancy)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 70, fill = "brown", color = "white", alpha = 0.7) +
  geom_density(color = "royalblue", fill = "lightblue", alpha = 0.6) +
  labs(
    title = "Distribution of Life Expectancy by Status",
    x = "Life Expectancy", y = "Density"
  ) +
  theme_minimal() +
  facet_wrap(~Status)

@@

2. **Diphtheria Immunization Coverage (`Diphtheria`)**: This variable represents the percentage of one-year-olds who have received the Diphtheria, Tetanus, and Pertussis (DTP3) immunization. Immunization coverage(@fig-diphtheria) is an important health indicator, as it reflects the effectiveness of a country's healthcare system in preventing infectious diseases, which in turn affects life expectancy.

@@

4. **Medical Doctors (`pct_doctor`)**: This variable indicates the number of medical doctors per 10,000 people (@fig-doctor). Access to healthcare professionals is a key determinant of health outcomes, and a higher number of doctors generally corresponds with better healthcare services and longer life expectancy.

5. **Nursing and Midwifery Personnel (`pct_nursing`)**: Similar to the number of doctors, this variable represents the number of nursing and midwifery personnel per 10,000 population (@fig-nursing). Nurses play a crucial role in delivering healthcare, and higher availability of nursing staff can contribute to improved health outcomes and increased life expectancy.

7. **Health Expenditure (`TotalExpenditure`)**: This variable reflects the general government expenditure on health as a percentage of total government expenditure, as shown in @fig-totalExpenditure. Higher health spending is often associated with better healthcare infrastructure, which can lead to longer life expectancy due to better disease prevention and treatment options.

```{r}
#| label: fig-status
#| fig-cap: "Histogram of Status, showing the counts of developed and developing countries in the dataset. There are 305 data points from developed countries and 788 from developing countries."
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(x = Status, fill = Status)) +
  geom_bar(stat = "count") +
  geom_text(stat = "count", aes(label = ..count..), vjust = 1.5, size = 5) +  # Move text to the bottom
  labs(x = "Country Status", 
       y = "Count", 
       title = "Distribution of Countries by Status (Developed vs Developing)") +
  scale_fill_manual(values = c("royalblue", "orange")) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| label: fig-diphtheria
#| fig-cap: "The plot shows similar immunization coverage conditions across developed and developing countries, with a noticeable increase in variance for developing countries due to some countries exhibiting low diphtheria immunization rates."
#| echo: false
#| warning: false
#| message: false

# Calculate variance for each Status group
variance_data <- data %>%
  group_by(Status) %>%
  summarise(variance = var(Diphtheria, na.rm = TRUE))

# Plot
ggplot(data, aes(x = Diphtheria, y = Status, fill = Status)) +
  geom_density_ridges(alpha = 0.7) +
  # Add text for Variance on the left side
  geom_text(data = variance_data, 
            aes(x = 5, y = Status, label = paste("Variance:", round(variance, 1))),
            size = 3, 
            hjust = 0, 
            color = "brown") +
  labs(x = "Diphtheria Immunization Coverage (%)", 
       y = "Country Status", 
       title = "Distribution of Diphtheria Immunization by Country Status") +
  scale_fill_manual(values = c("royalblue", "darkorange")) +
  scale_x_continuous(limits = c(0, 100)) +  # Ensure x-axis is between 0 and 100
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| label: fig-schooling
#| fig-cap: "Violin plot with boxplot for Schooling by Status, highlighting a narrow distribution for developed regions with a higher median of 16 years of schooling. In contrast, developing countries show a wider variance and a relatively lower median of around 12.5 years, indicating significant disparities in educational access and attainment, with developed countries having more consistent and higher levels of education compared to the varied and generally lower levels in developing countries."
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(
  x = Status,
  y = Schooling,
  fill = Status
)) +
  geom_violin(alpha = 0.5, position = position_dodge(width = 1)) + 
  geom_boxplot(width = 0.1, color = "white", outlier.shape = NA, position = position_dodge(width = 1)) +
  labs(
    x = "Status",
    y = "Schooling Years",
    title = "Distribution of Schooling Years by Status"
  ) +
  scale_fill_manual(values = c("Developed" = "brown", "Developing" = "darkgreen")) + 
  # Separate panels for each status
  facet_wrap(~Status, scales = "free_x") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r}
#| label: fig-totalExpenditure
#| fig-cap: "Boxplot for government expenditure on health as a percentage of total expenditure, showing relatively high percentages for developed countries compared to developing countries, highlighting that developed countries' governments typically allocate more resources to health, though there are exceptions."
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(x = Status, y = TotalExpenditure, fill = Status)) +
  # Adding boxplots on top of violin plots for more detail on median and quartiles
  geom_boxplot(width = 0.5, alpha = 0.7, position = position_dodge(0.9)) +

  labs(
    title = "Total Government Health Expenditure by Status",
    x = "Country Status",
    y = "Total Government Expenditure on Health (%)"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("skyblue", "orange")) +  # Customize colors for Developed and Developing
  theme(legend.position = "none")  # Hide the legend for clarity

@@

```{r}
#| label: fig-doctor
#| fig-cap: "Density plot for doctors per 10,000 population by Status, showing that the density of medical doctors in developed regions is relatively higher compared to developing regions. In developing countries, the density is concentrated at lower levels, suggesting that the healthcare system is generally stronger and more accessible in developed countries."
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(x = pct_doctor, y = Status, fill = Status)) +
  geom_density_ridges(alpha = 0.7) +
  labs(x = "Density of Medical Doctor", y = "Country Status", title = "Medical Doctor Condition") +
  scale_fill_manual(values = c("royalblue", "darkorange")) +
  theme_minimal() +
  theme(legend.position = "none")

@@

```{r}
#| label: fig-nursing
#| fig-cap: "Number of Nurses by Status with Highlighted Mean and Dotplot. The mean for developed countries is around 80, while the mean for developing countries is around 35, with many data points showing low nurse density in developing countries. Both status groups have medians lower than their means, suggesting a right-skewed distribution with a concentration of countries with fewer nurses."
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(x = Status, y = pct_nursing, fill = Status)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +  # Boxplot for the distribution
  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.5, alpha = 0.4, position = "dodge") +  # Smaller dots with dodge position
  stat_summary(fun = "mean", geom = "point", shape = 18, size = 4, color = "black", fill = "yellow") +  # Highlight the mean
  labs(
    title = "Number of Nurses by Status with Highlighted Mean and Dotplot",
    x = "Status",
    y = "Number of Nurses per 10000"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("brown", "darkgreen")) +  # Customize fill colors
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Tilt x-axis labels

@@

The models assume a linear relationship between the predictors and life expectancy, where each predictor contributes independently or interactively to the variation in the outcome variable. Define $y_i$ as the Life Expectancy for country $i$. The factors $x_1$, $x_2$,...represent socioeconomic predictors such as Diphtheria, Schooling, TotalExpenditure, pct_doctor, pct_nursing, and interaction term.

@@

Linear regression assumes that the residuals errors of the model are normally distributed. If this assumption is violated, it could lead to biased estimates and incorrect conclusions. This issue is especially important in cross-country data, where life expectancy may be influenced by diverse factors that do not follow a normal distribution.

Predictors such as the number of doctors per 10,000 and the number of nurses per 10,000 might be correlated with each other. High correlation between predictors can cause multicollinearity, making it difficult to assess the individual contribution of each predictor.

@@

In @tbl-keyestimates, the Developed model is more efficient and better fitted primarily due to its higher R² (64.5% vs 51.5%), indicating that it explains a greater proportion of variance in life expectancy. Additionally, its RMSE value is lower (3.78 vs 5.06), showing more precise predictions. The Developing model has a larger sample size and variability, which might have diluted the effect of individual predictors. The better fit of the Developed model is partly due to the smaller, more homogeneous sample, allowing it to provide more reliable estimates with fewer outliers.

### Model for Developed Countries

For the Developed model, the predictors considered included Diphtheria Immunization, Schooling, Total Expenditure, and Number of Doctors per 10000. The coefficients and standard errors for these predictors were as follows:

**Diphtheria Immunization**: Coefficient = 0.022, Standard Error = 0.025

**Schooling**: Coefficient = 0.801, Standard Error = 0.151

**Health Expenditure**: Coefficient = 0.056, Standard Error = 0.084

**Number of Medical Doctors per 10000**: Coefficient = 0.018, Standard Error = 0.034

These coefficients indicate that `Schooling` has the highest contribution to explaining life expectancy, followed by `Diphtheria Immunization`. The relatively small contribution of `#Doctors per 10000` suggests that doctor practices may have a less direct impact on life expectancy in developed countries, but this could be nuanced by other health factors not captured in the model.

### Model for Developing Countries

For the Developing model, the same predictors were considered, but their relationships with life expectancy differed. The coefficients and standard errors for each predictor were as follows:

**Diphtheria Immunization**: Coefficient = 0.036, Standard Error = 0.010

**Schooling**: Coefficient = 1.727, Standard Error = 0.113

**Number of Medical Doctors per 10000**: Coefficient = 0.090, Standard Error = 0.021

**Number of Nurses per 10000**: Coefficient = -0.003, Standard Error = 0.003

In the Developing model, `Schooling` again has the most significant impact, and `#Doctors per 10000` is also a substantial predictor. This highlights the importance of healthcare workforce availability in improving life expectancy in developing countries, unlike the developed context where Schooling was more prominent. The negative relationship of `Total Expenditure` suggests that in developing countries, spending in certain healthcare sectors could be misallocated or inefficient.
```{r}
#| label: tbl-keyestimates
#| tbl-cap: "Summary of key model estimates for Developed and Developing counties, including coefficients for predictors like Diphtheria, Schooling, and TotalExpenditure, pct_doctor, pct_nursing, with standard errors for each estimate. Model performance statistics, such as sample size, R², and adjusted R², are also displayed."
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| echo: false
#| label: tbl-modelresults
#| tbl-cap: "The prediction table indicates that the average life expectancy in developed countries is 78.53 years, while in developing countries, it is slightly lower at 71.68 years. This difference of approximately 5.1 years in life expectancy suggests a significant health and infrastructure disparity between developed and developing countries. These results align with broader global trends where developed countries typically benefit from better healthcare systems, higher standards of living, and advanced medical technology, contributing to longer life expectancies. "
#| eval: true
#| warning: false
#| message: false

# load data and model
data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))
lm_developed <- readRDS(here::here("models/lm_developed.rds"))
lm_developing <- readRDS(here::here("models/lm_developing.rds"))

# display avg percent and normalized percent
results %>%
  select(AverageDeveloped, AverageDeveloping) %>%
  kable(
    col.names = c("Average Life Expectancy in Developed Countries", "Average Life Expectancy in Developing Countries"),
    digits = 2,
    booktabs = TRUE
  ) %>%
  kable_styling(
    latex_options = c("striped", "scale_down")
  )
```

@@

data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))
lm_developed <- readRDS(here::here("models/lm_developed.rds"))
lm_developing <- readRDS(here::here("models/lm_developing.rds"))

# Make predictions
data <- data %>%
  mutate(
    PredictedDeveloped = predict(lm_developed, newdata = data),
    PredictedDeveloping = predict(lm_developing, newdata = data)
  )

@@

>>>>

---
title: "What to Expect with Life Expectancy?"
subtitle: "Examining How Healthcare Expenditure, Education, and Medical Infrastructure Influence Life Expectancy Across Developed and Developing Regions"
author: 
  - Manjun Zhu
thanks: "Code and data are available at: [https://github.com/Karrrrmen/Life_Expectancy](https://github.com/Karrrrmen/Life_Expectancy)."
date: today
date-format: long
abstract: "Life expectancy has seen unprecedented growth in recent decades, driven by advancements in healthcare, improved access to medical services, and reduced child mortality. However, this growth has not been evenly distributed across the globe, with wealthier nations benefiting disproportionately. This research examines the impact of a nation’s educational standing and health conditions on life expectancy, exploring potential variables beyond those traditionally studied. By developing a linear model, the analysis identifies key relationships between education and healthcare indicators and their influence on population longevity. The findings aim to inform health policy and education planning, providing actionable insights to address disparities and promote equitable growth in life expectancy worldwide."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

# load libraries
library(dplyr)
library(tidyverse)
library(here)
library(arrow)
library(ggplot2)
library(knitr)
library(kableExtra)
library(modelsummary)
library(ggbeeswarm)
library(ggridges)
library(broom)

# load data
data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))

@@

Life expectancy has witnessed remarkable growth over the past century, driven by advancements in healthcare, improved access to medical services, and reduced child mortality rates. However, this progress has not been distributed equitably across the globe. While nations with robust economies and developed healthcare systems have achieved significant gains, others, particularly in underdeveloped regions, continue to face barriers such as inadequate infrastructure, persistent diseases, and economic instability. Despite existing research on life expectancy determinants, gaps remain in understanding the nuanced relationships between economic standing, healthcare access, and other critical variables.

This study aims to examine the factors influencing life expectancy across different nations by developing a linear regression model that incorporates both healthcare and education variables. The response variable in the analysis is life expectancy, while the predictors include national status (Developed vs. Developing), diphtheria immunization coverage, government healthcare expenditure as a percentage of total expenditure, and the number of doctors and nurses per 10,000 people. Additionally, schooling is considered as an important educational factor. By quantifying the effects of these predictors, the study seeks to shed light on disparities in life expectancy between developed and developing nations, and to identify key factors contributing to these differences. Through this analysis, we aim to contribute valuable insights for policy development focused on improving healthcare and economic outcomes.

Previous studies provide valuable context for this research. One study [@Garon2015] examined the long-term effects of diphtheria on survivors, revealing that paralyzed individuals experience greater reductions in life expectancy, which underscores the importance of healthcare infrastructure in mitigating preventable diseases in underdeveloped regions. Another study extended this understanding by exploring the role of the Human Development Index (HDI), emphasizing the influence of public health initiatives, economic conditions, and education on life expectancy in low and medium-HDI nations [@Girum2018]. Besides that, developed econometric techniques, including panel cointegration analysis, have demonstrated a positive relationship between health care expenditure per capita and life expectancy, providing a robust foundation for integrating economic variables into this study [@He2020]. Additionally, a finding [@Hanushek2016] underscored the importance of combining both educational investments and government support in improving life expectancy, particularly in developing regions. The strong correlation between education and health suggests that higher levels of education improve overall health, reducing mortality rates and extending life expectancy.

The analysis reveals that life expectancy in developed countries is generally higher and more stable, attributed to stronger healthcare systems, consistent access to medical services, and robust economic stability. These factors are typically supported by well-established public health policies. In contrast, life expectancy in developing countries demonstrates a broader range of variation, though with a noticeable upward trend. This improvement is driven by diverse factors like better healthcare access, economic development, and increasing resource availability, but the inconsistency across regions highlights disparities in infrastructure. The faster growth trajectory in these nations indicates progress, yet challenges remain in achieving stable health outcomes. These findings suggest that policymakers in developing countries should prioritize healthcare system strengthening, resource access, and reducing regional inequalities, while in developed countries, maintaining consistent healthcare access and supporting sustained economic growth should remain priorities.

@@

This study utilizes two primary datasets related to global health and life expectancy. The first dataset, Life Expectancy and Healthy Life Expectancy [@world_health_statistics], provides comprehensive global health data, including variables such as GDP, life expectancy, mortality rates, and access to healthcare. The second dataset, Life Expectancy (WHO) [@life_expectancy_who], focuses specifically on life expectancy trends and factors, offering a detailed breakdown by country, year, and socioeconomic variables.

@@

We performed data cleaning and analysis using R [@citeR], incorporating a range of packages for efficient data manipulation, visualization, and statistical modeling, leveraging packages such as **`tidyverse`** for data manipulation and visualization [@tidyverse], **`dplyr`** for streamlined data wrangling [@dplyr], **`ggplot2`** for creating high-quality plots [@ggplot2], **`ggridges`** for displaying the distribution of variables across different categories with ridge plots [@ggridges], **`ggbeeswarm`** for visualizing the distribution of data points with swarming plots [@ggbeeswarm], **`here`** for managing file paths [@here], **`lubridate`** for handling date-related variables [@lubridate], **`arrow`** for data storage and processing [@arrow], **`kableExtra`** for generating polished tables [@kableExtra], **`modelsummary`** for summarizing model results [@modelsummary],  **`broom`** for tidying up model outputs into easily readable data frames [@broom], and **`knitr`** for dynamic report generation and the integration of code with results in reproducible workflows [@knitr].

@@

The measurement process explains how real-world phenomena—such as healthcare access, economic indicators, or demographic factors—are translated into structured data entries. Each row in the dataset represents a specific country in a given year, capturing various variables related to life expectancy and its determinants. These variables originate from data collection efforts by global organizations like the World Health Organization (WHO), which rely on surveys, national health systems, and standardized reporting frameworks to compile accurate and comparable statistics across countries. Below is a description of key variables in the dataset:

@@

**Percentage of Health Expenditure (`HealthExpenditure`)**: This represents the general government expenditure on health as a percentage of total government expenditure. It is reported by national governments or international organizations and recorded as a percentage in the dataset.

@@

```{r}
#| label: fig-response
#| fig-cap: "The distribution highlights the frequency of life expectancy values across the dataset. The distribution is slightly right-skewed, with a mean around 73 years, reflecting the overall health trends captured in the data."
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(x = LifeExpectancy)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 70, fill = "#433769", color = "white", alpha = 0.7) +
  geom_density(color = "#8980a8", fill = "#d0c7eb", alpha = 0.6) +
  labs(
    title = "Distribution of Life Expectancy",
    x = "Life Expectancy", y = "Frequency"
  ) +
  theme_minimal()

@@

2. **Diphtheria Immunization Coverage (`Diphtheria`)**: This variable represents the percentage of one-year-olds who have received the Diphtheria, Tetanus, and Pertussis (DTP3) immunization. Immunization coverage[@fig-diphtheria] is an important health indicator, as it reflects the effectiveness of a country's healthcare system in preventing infectious diseases, which in turn affects life expectancy.

@@

4. **Medical Doctors (`pct_doctor`)**: This variable indicates the number of medical doctors per 10,000 people[@fig-doctor]. Access to healthcare professionals is a key determinant of health outcomes, and a higher number of doctors generally corresponds with better healthcare services and longer life expectancy.

5. **Nursing and Midwifery Personnel (`pct_nursing`)**: Similar to the number of doctors, this variable represents the number of nursing and midwifery personnel per 10,000 population[@fig-nursing]. Nurses play a significant role in delivering healthcare, and higher availability of nursing staff can contribute to improved health outcomes and increased life expectancy.

7. **Percentage of Health Expenditure (`HealthExpenditure`)**: This variable reflects the general government expenditure on health as a percentage of total government expenditure, as shown in @fig-totalExpenditure. Higher health spending is often associated with better healthcare infrastructure, which can lead to longer life expectancy due to better disease prevention and treatment options.

```{r}
#| label: fig-status
#| fig-cap: "The distribution of Status shows the counts of developed and developing countries in the dataset. There are 305 data points from developed countries and 788 from developing countries."
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(x = Status, fill = Status)) +
  geom_bar(stat = "count") +
  labs(x = "Country Status", 
       y = "Frequency", 
       title = "Distribution of Countries by Status (Developed vs Developing)") +
  scale_fill_manual(values = c("#b3685d", "#a85b6a")) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| label: fig-diphtheria
#| fig-cap: "The distribution of immunization coverage for diphtheria reveals that immunization coverage is generally high, with most countries achieving high vaccination rates. However, there is some variance, reflecting disparities in immunization efforts across different regions."
#| echo: false
#| warning: false
#| message: false

data_clean <- data %>%
  filter(!is.na(Diphtheria) & Diphtheria > 0)

ggplot(data_clean, aes(x = Diphtheria)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 70, fill = "#4d1f0a", color = "white", alpha = 0.7) +
  labs(
    title = "Distribution of Immunization Coverage of Diphtheria",
    x = "Immunization Coverage of Diphtheria", y = "Frequency"
  ) +
  theme_minimal()

@@

```{r}
#| label: fig-schooling
#| fig-cap: "The distribution of Schooling Years is shown as a representation of the Education Factor, the distribution of schooling years across the dataset approaches a normal distribution"
#| echo: false
#| warning: false
#| message: false

data_clean <- data %>%
  filter(!is.na(Schooling) & Schooling > 0)

ggplot(data_clean, aes(x = Schooling)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 15, fill = "#629596", color = "white", alpha = 0.7) +
  labs(
    title = "Distribution of Schooling Years",
    x = "Schooling Years", y = "Frequency"
  ) +
  theme_minimal()

@@

```{r}
#| label: fig-totalExpenditure
#| fig-cap: "The distribution of government expenditure on health as a percentage of total expenditure reflects the level of medical support provided by governments."
#| echo: false
#| warning: false
#| message: false

data_clean <- data %>%
  filter(!is.na(TotalExpenditure) & TotalExpenditure > 0)

ggplot(data_clean, aes(x = TotalExpenditure)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 20, fill = "#5765ad", color = "white", alpha = 0.7) +
  labs(
    title = "Distribution of Government Expenditure on Health",
    x = "Percentage of Health Expenditure as a Percentage of Total Expenditure", y = "Frequency"
  ) +
  theme_minimal()

@@

```{r}
#| label: fig-doctor
#| fig-cap: "The distribution of doctors per 10,000 population, providing an overview of the distribution of medical doctors globally, reflecting general medical resource availability."
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(x = pct_doctor)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 15, fill = "#8b6fb0", color = "white", alpha = 0.7) +
  labs(
    title = "Distribution of the number of doctors per 10,000",
    x = "Number of medical doctors per 10,000 population", 
    y = "Frequency"
  ) +
  theme_minimal()

```{r}
#| label: fig-nursing
#| fig-cap: "The distribution of nurses per 10,000 population, offering a broad view of nurse density worldwide as an indicator of healthcare support."
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(x = pct_nursing)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 15, fill = "#3f594a", color = "white", alpha = 0.8) +
  labs(
    title = "Distribution of the number of nurses per 10000",
    x = "the number of nurses and  per 10000 population", y = "Frequency"
  ) +
  theme_minimal()

@@

The models assume a linear relationship between the predictors and life expectancy, where each predictor contributes independently or interactively to the variation in the outcome variable. Define $y_i$ as the Life Expectancy for country $i$. The factors $x_1$, $x_2$,...represent socioeconomic predictors such as Diphtheria Immunization Coverage, Schooling, Health Expenditure, Number of Doctors and Nurses per 10000, and interaction term.

@@

The linear regression model used to estimate life expectancy relies on several key assumptions that must be met for valid results. First, it assumes linearity, meaning there is a straight-line relationship between the predictors (e.g., healthcare condition, education levels) and the outcome variable (`life expectancy`). This assumption ensures that a change in a predictor corresponds to a consistent change in life expectancy. Second, the model assumes that the errors are uncorrelated, meaning the residuals do not exhibit any patterns when plotted. This assumption is significant for making accurate inferences, such as significance tests and confidence intervals. Third, the model assumes homoscedasticity, which means that the variance of the residuals should remain constant across all values of the predictors. This assumption helps avoid bias in estimating the relationship between the predictors and the outcome variable. Fourth, the residuals should follow a normal distribution. This assumption is necessary for performing accurate hypothesis tests and for generating valid confidence intervals, which provide insights into the uncertainty of the model's predictions. Lastly, it is important that the model does not appear multicollinearity, where predictors are highly correlated with each other. These assumptions provide a foundation for the model, violations can lead to misleading results with inaccurate coefficients.

However, several limitations apply to this model. Despite the assumption of linearity, real-world relationships between life expectancy and its predictors may be more complex or non-linear, and such complexity would not be fully captured by this model. For instance, the effect of economic development on life expectancy may vary significantly at different levels of income. Moreover, while the model assumes uncorrelated errors, there may be factors that influence both the predictors and life expectancy, leading to correlated residuals. For example, social determinants like income inequality or access to healthcare might impact both education and life expectancy, potentially violating the assumption of independence. The assumption of constant variance also has limitations, as certain groups (e.g., countries with lower health expenditure) may have more variability in life expectancy than others, potentially leading to not constant variance. Additionally, the assumption of normally distributed errors may not hold in all cases, especially when extreme outliers (such as countries with unusual life expectancy rates) are present in the data.

Other external factors not accounted for in this model may also influence life expectancy. Environmental factors like climate change, sudden public health crises (e.g., pandemics), and political instability could have significant effects on life expectancy, yet these factors may not be included as predictors in the model. These limitations should be considered when interpreting the model’s results, particularly when applying the findings to specific populations or making broad predictions.

@@

In @tbl-keyestimates, the Developed model is more efficient and better fitted primarily due to its higher R² (64.5% vs 51.5%), indicating that it explains a greater proportion of variance in life expectancy. Additionally, its RMSE value is lower (3.76 vs 4.94), showing more precise predictions. The Developing model has a larger sample size and variability, which might have diluted the effect of individual predictors. The better fit of the Developed model is partly due to the smaller, more homogeneous sample, allowing it to provide more reliable estimates with fewer outliers.

Extended versions of these tables can be found in @tbl-additionaltable

```{r}
#| echo: false
#| label: tbl-modelresults
#| tbl-cap: "The prediction table indicates that the average life expectancy in developed countries is 78.52 years, while in developing countries, it is slightly lower at 71.67 years. This difference of approximately 5.1 years in life expectancy suggests a significant health and infrastructure disparity between developed and developing countries. These results align with broader global trends where developed countries typically benefit from better healthcare systems, higher standards of living, and developed medical technology, contributing to longer life expectancies. "
#| eval: true
#| warning: false
#| message: false

# display avg percent and normalized percent
results %>%
  select(AverageDeveloped, AverageDeveloping) %>%
  kable(
    col.names = c("Average Life Expectancy in Developed Countries", "Average Life Expectancy in Developing Countries"),
    digits = 2,
    booktabs = TRUE
  ) %>%
  kable_styling(
    latex_options = c("striped", "scale_down")
  )

```{r}
#| label: tbl-keyestimates
#| tbl-cap: "Summary of key model estimates for Developed and Developing counties, including coefficients for predictors like Diphtheria, Schooling, and TotalExpenditure, pct_doctor, pct_nursing, with standard errors for each estimate. Model performance statistics, such as sample size, R², and adjusted R², are also displayed."
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

@@

lm_developed <- readRDS(here::here("models/lm_developed.rds"))
lm_developing <- readRDS(here::here("models/lm_developing.rds"))

# Make predictions
data <- data %>%
  mutate(
    PredictedDeveloped = predict(lm_developed, newdata = data),
    PredictedDeveloping = predict(lm_developing, newdata = data),
  )

@@

The predicted life expectancy distributions across various predictors reveal important trends. Longer schooling years[@fig-schoolgroups] strongly correlate with higher predicted life expectancy, emphasizing the positive impact of education. High diphtheria immunization coverage[@fig-diphtheria-groups] is also associated with increased life expectancy, though this relationship is less consistent for medium and low coverage levels, indicating variability in its predictive contribution. Similarly, higher government health expenditure[@fig-totalexpenditure-groups] aligns with greater life expectancy, while lower expenditure shows limited variability in predicted outcomes. Lastly, the distribution of medical professionals[@fig-doctornursing-groups] highlights that developed medical conditions generally contribute to higher life expectancy, though some variability in their influence remains.

```{r}
#| echo: false
#| eval: true

# Load the model
lm_overall <- readRDS(here::here("models/lm_overall.rds"))

# Generate predictions
prediction_data <- data %>%
  filter(!is.na(Diphtheria) & !is.na(TotalExpenditure) & !is.na(Schooling) & !is.na(pct_doctor) & !is.na(pct_nursing))

prediction_data$predictedLifeExpectancy <- predict(lm_overall, newdata = prediction_data)

```{r}
#| echo: false
#| eval: true
#| label: fig-schoolgroups
#| fig-cap: "The distribution shows how different levels of education correlate with predicted life expectancy, highlighting group-wise distribution and variability."
#| warning: false
#| message: false

# Remove rows with 0 or NULL values in Schooling
prediction_data <- prediction_data %>%
  filter(!is.na(Schooling) & Schooling > 0)

# Create Schooling year groups for segmentation
prediction_data$schooling_years <- cut(
  prediction_data$Schooling,
  breaks = c(3.1, 6.1, 12.1, 15.1, 18.1, 21),
  labels = c("3-6", "6-12", "12-15", "15-18", "18-21")
)
